Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2020-04-01 anzjs_elections.Rmd
Results and Conclusion changed to past tense - at the request of the formatting/technical editor

The spatial autoregressive coefficient $\rho$ is positive and significant in the 2001, 2004 and 2016 elections (Figure \@ref(fig:rhovis)). In these three elections, there is evidence to suggest that neighbours share some influential characteristics outside the explanatory variables, which affect the two-party preferred vote. Conversely, in the other three elections, the spatial effect weakens to become insignificant (although still positive).

@@

Since all socio-demographics have been standardised to have a mean of zero and a variance of one, the intercept in each model can be interpreted as the estimated two-party preferred vote for an electorate with mean characteristics (aside from `Judaism`, `Indigenous`, `Islam` and `Buddhism`, where it assumes the mean of the log value). Figure \@ref(fig:plotintercept) shows that the baseline of party preference has varied over the elections, with the biggest swing occurring in the 2007 election where the mean electorate shifted more than five percentage points in favour of the Labor party.

@@

(ref:figcappartresplot1) Partial residual plots by election year for a selection of predictors. Linear model with 95% confidence bands overlaid. Most predictors have a positive relationship: the larger the value the more likely the electorate preferences Liberal. The relationship is relatively robust over time, with the exception of `DiffAddress`, `Incomes`, `OtherLanguageHome` and `Pop_00_19`.

@@

(ref:figcappartresplot2) Partial residual plots by election year for a selection of predictors. Linear model with 95% confidence bands overlaid. Several predictors have a negative relationship: with larger values indicating the electorate more likely preferences Labor. Most relationships are relatively stable over elections, except `Unemployment` and `Education`."

@@

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Partial residuals, for a given variable, are the residuals from the fitted model with the estimated effect of that variable added to it. These plots show the direction, size and significance of an estimated effect, as well as any deviations from linearity. In each plot, the slope of the prediction line matches the estimated coefficient and the shaded region represents a 95% confidence band. Plots are computed using the method in @visreg. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. The estimated intercept is also added to the partial residuals for interpretability. Plots for each election are faceted to compare the effects over time in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Only socio-demographics that have a significant effect in at least two elections are displayed in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2).

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) are consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 1 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) are also more likely to support the Liberal party (see row 3 in Figure \@ref(fig:partresplot1)), from 2004 onwards.

Similarly, the proportion of workers in managerial, administrative, clerical and sales roles (`ManagerAdmin`), is also a significant predictor of two-party preference vote across all six elections, with a higher proportion of people working these jobs increasing Liberal support.

Of these job related variables, the most surprising effect is that associated with the proportion of workers in education, healthcare, social work, community and arts (`SocialServ`). Typically the Labor party has more generous funding schemes affecting these areas of work, so one might expect `SocialServ` to have a negative effect on two-party preference. However, in every election this effect is found to be positive and significant.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appear more likely to support the Liberal party, as the `Incomes` factor has a positive effect on Liberal preference (see row 6 in Figure \@ref(fig:partresplot1)). This effect is significant in every election aside from 2004 and 2010. `Unemployment` however, is not as influential. In 2001 and 2004, electorates with higher unemployment align with Labor, but over time this shifts towards support for the Liberal party, culminating in a positive (insignificant) effect in 2016.

@@

The older Australian population is often believed to be more conservative, and the left leaning political parties (including Labor) typically have a stronger appeal to younger people. This effect is indeed observed across the six elections, with populations between 20 and 34 years of age (`Pop_20_34`) being very strongly aligned with Labor preference (bottom row in Figure \@ref(fig:partresplot2)). Larger populations of 35 to 54 year olds (`Pop_35_54`) are also associated with Labor, but the magnitude of this effect is far smaller. Populations under 20 years of age is only significant in 2007, where `Pop_00_19` increased Liberal support.

@@

Since 2007, electorates with higher education levels are associated with supporting the Labor party, with this effect being significant in 2007, 2013 and 2016 and only marginally insignificant in 2010. In the elections before 2007, education has a negligible effect (see row 3 in Figure \@ref(fig:partresplot2)). Additionally, student populations (`CurrentlyStudying`) do not affect electoral party preference in any election (not shown).

@@

Larger migrant populations from Asia, the Middle East, South-Eastern Europe, the United Kingdom and elsewhere, are either associated with Labor support, or have no effect. Of these areas, only South-Eastern European populations significantly affect party preference, with larger populations associating with Labor in 2013 and 2016 (row 2, Figure \@ref(fig:partresplot2)). Speaking other languages (aside from English) however, appears to have a far stronger effect, as observed through the `OtherLanguage` variable. Electorates with more diverse speech are associated with higher support for the Liberal party from 2004 onwards, with this effect being significant in 2013 and 2016 (see row 7, Figure \@ref(fig:partresplot1)). Furthermore, none of the variables relating to religious beliefs aside from Christianity have a material effect in any election (this includes the Buddhist, Muslim, Jewish, non-religious and Indigenous Australian populations). The association between Christian populations (`Christianity`) and the Liberal party steadily increases over the years, becoming positive and significant in 2016.

@@

In 2001, 2004 and 2016, higher proportions of people that have recently (in the past five years) moved house (`DiffAddress`) increased electoral support for the Liberal party (see row 5 in Figure \@ref(fig:partresplot1)). This is somewhat surprising as one might expect house ownership and rental prices to be linked to two-party preference, rather than household mobility (`PropertyOwned` and `RentLoan` are not significant in any election).

Higher proportions of single parent households are associated with Labor support in all elections (albeit insignificant in 2004, see row 6 in Figure \@ref(fig:partresplot2)), whereas the electoral family and household sizes (via the `FamHouseSize` variable) do not appear to be associated with either party.

@@

From 2007 onwards, both marriages (`Married`) and de facto relationships (`DeFacto`) are found to be strong predictors of the two-party preferred vote in favour of the Labor party. In 2001 and 2004 neither of these variables are significant (see rows 4 and 5 in Figure \@ref(fig:partresplot2)).

@@

It is often hypothesised that states have a systematic bias towards one of the two major parties. Boxplots of residuals grouped by state (Figure \@ref(fig:resstate)) show that the data reflects this to only a limited extent. Tasmania and the Australian Capital Territory appear to have a bias towards Labor, whereas the South Australia and the Northern Territory tend towards voting Liberal. However, there are relatively few electorates in each of these states (five, two, eleven and two respectively), so this apparent result may be due to incumbent effects rather than an actual state-specific bias.

@@

The incumbent party appears to have a distinct advantage at the next election. The boxplots in Figure \@ref(fig:resincumbent) show that if either of the Labor or Liberal parties won the seat at the previous election, the electorate is likely to vote in their favour, over and above any socio-demographic effects---this effect has not been captured by the model.

@@

Three robustness checks are conducted to confirm model stability. First, a model for each election is re-estimated using only the variables that are found to be significant in at least one of the six elections. The estimated coefficients of the variables in the re-estimated models all fall within their respective 95% confidence intervals from the full models. The second check involves the ten largest pairwise correlations. For each pair, a model for each election is re-estimated omitting one of the two variables. It is found that for each of these pairs, the estimated effect of the remaining variable in the reduced model lies within the 95% confidence interval from the full model. The final check is a visual exploration of different variable projections using a tour [@tourr] for each election. No definitive signs of multicollinearity are observed, and as expected (given the nature of spatial data), there is some clumping of electorates for certain projections.

@@

Based on the distribution of the Cook's distance values and the distribution of hat values, a Cook's distance greater than $0.1$ is considered to be influential, and a hat value greater than $0.5$ is considered to have high leverage. Electorates fitting these criteria are flagged and investigated to examine the characteristics driving these values.

The electorate of Sydney (NSW) has a large Cook's distance and high leverage from 2001 to 2007, due to its diverse population (languages, birthplace and religion), high density of young adults (20 to 34 years old), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and the extent of this support is underpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup---implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections since, the model underpredicts Wentworth's Liberal support.

Lingiari, an electorate taking up almost all of the Northern Territory, has consistently high leverage (all years) and is an outlier in all but the 2013 election due to its large Indigenous population, low rates of property ownership and few workers in management or administrative jobs. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, as well as a high proportion of single parent households. These characteristics explain its high leverage in 2001, 2004, 2010 and 2013, and its strong Labor support makes it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance are Canberra (ACT) and Durack (WA) in 2013, and Solomon (NT) in 2016.

All of the electorates examined are not unduly influential in the model and therefore no action is required.

@@

This paper explores the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and censuses. As a census does not always occur in the same year as an election, census data for the 2004--2013 elections are generated by employing a method of spatio-temporal imputation. This imputes electoral socio-demographics for the electoral boundaries in place at the time of the election---an approach that is distinctly different from previous work on modelling election outcomes, where census and election data are typically joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables are standardised (to adjust for changing variable scales) and subsets of variables (representing similar information) are combined into factors, resulting in a reduced predictor set. A spatial error model is then estimated for each election, accounting for the inherent spatial structure of the data.

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote are found to remain steady, whilst a few (typically weaker) effects vary over time. Industry and type of work are particularly influential. Energy-related and manufacturing/construction jobs, as well as administrative roles and jobs in education and social services are strongly linked with the Liberal party in all elections. Incomes have a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shift from weak association with Labor to a significant Liberal effect over the years, and higher education levels are associated with Labor from 2007 (although marginally insignificant in 2010). Electorates with large populations 20 to 34 years are strongly associated with Labor, whilst the 35 to 54 year old bracket also increases Labor support, but to a lesser extent. It is also found that birthplace diversity slightly favours Labor, relationships (both marriages and de facto relationships) align with Labor preference from 2010 onwards, and the influence of Christian populations has trended towards Liberal support whilst other religions have negligible effects. Family and household sizes have minimal influence, although electorates with more single parent households are linked with Labor support. Furthermore, the spatial effects are found to be positive in all elections and significant in 2001, 2004 and 2016, meaning that other characteristics that neighbours have in common  (outside of the variables in the model) appear to be influential in those years.

>>>>

The spatial autoregressive coefficient $\rho$ was positive and significant in the 2001, 2004 and 2016 elections (Figure \@ref(fig:rhovis)). In these three elections, there is evidence to suggest that neighbours shared some influential characteristics outside the explanatory variables, which affected the two-party preferred vote. Conversely, in the other three elections, the spatial effect is weaker and insignificant (although still positive).

@@

Since all socio-demographics were standardised to have a mean of zero and a variance of one, the intercept in each model can be interpreted as the estimated two-party preferred vote for an electorate with mean characteristics (aside from `Judaism`, `Indigenous`, `Islam` and `Buddhism`, where it assumes the mean of the log value). Figure \@ref(fig:plotintercept) shows that the baseline of party preference varied over the elections, with the biggest swing occurring in the 2007 election where the mean electorate shifted more than five percentage points in favour of the Labor party.

@@

(ref:figcappartresplot1) Partial residual plots by election year for a selection of predictors. Linear model with 95% confidence bands overlaid. Most predictors had a positive relationship: the larger the value the more likely the electorate preferenced Liberal. The relationships were relatively robust over time, with the exception of `DiffAddress`, `Incomes`, `OtherLanguageHome` and `Pop_00_19`.

@@

(ref:figcappartresplot2) Partial residual plots by election year for a selection of predictors. Linear model with 95% confidence bands overlaid. Several predictors had a negative relationship: with larger values indicating the electorate more likely preferenced Labor. Most relationships were relatively stable over elections, except `Unemployment` and `Education`."

@@

To investigate the socio-demographics that had a strong effect on the two-party preferred vote, partial residual plots are used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Partial residuals, for a given variable, are the residuals from the fitted model with the estimated effect of that variable added to it. These plots show the direction, size and significance of an estimated effect, as well as any deviations from linearity. In each plot, the slope of the prediction line matches the estimated coefficient and the shaded region represents a 95% confidence band. Plots were computed using the method in @visreg. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. The estimated intercept is also added to the partial residuals for interpretability. Plots for each election are faceted to compare the effects over time in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Only socio-demographics that had a significant effect in at least two elections are displayed in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2).

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) were consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 1 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party has maintained close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) were also more likely to support the Liberal party (see row 3 in Figure \@ref(fig:partresplot1)), from 2004 onwards.

Similarly, the proportion of workers in managerial, administrative, clerical and sales roles (`ManagerAdmin`), was also a significant predictor of two-party preference vote across all six elections, where higher proportions of people working these jobs increased Liberal support.

Of these job related variables, the most surprising effect is that associated with the proportion of workers in education, healthcare, social work, community and arts (`SocialServ`). Typically the Labor party has more generous funding schemes affecting these areas of work, so one might expect `SocialServ` to have had a negative effect on two-party preference. However, in every election this effect was found to be positive and significant.

@@

Typically the Labor party campaigns on more progressive policies, often including tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appeared more likely to support the Liberal party, as the `Incomes` factor had a positive effect on Liberal preference (see row 6 in Figure \@ref(fig:partresplot1)). This effect was significant in every election aside from 2004 and 2010. `Unemployment` however, was not as influential. In 2001 and 2004, electorates with higher unemployment aligned with Labor, but over time this shifted towards support for the Liberal party, culminating in a positive (insignificant) effect in 2016.

@@

The older Australian population is often believed to be more conservative, and the left leaning political parties (including Labor) typically have a stronger appeal to younger people. This effect is indeed observed across the six elections, where populations between 20 and 34 years of age (`Pop_20_34`) aligned strongly with Labor preference (bottom row in Figure \@ref(fig:partresplot2)). Larger populations of 35 to 54 year olds (`Pop_35_54`) were also associated with Labor, but the magnitude of this effect was far smaller. Populations under 20 years of age was only significant in 2007, where `Pop_00_19` increased Liberal support.

@@

From 2007, electorates with higher education levels have associated with supporting the Labor party, with this effect being significant in 2007, 2013 and 2016 and only marginally insignificant in 2010. In the elections before 2007, education had a negligible effect (see row 3 in Figure \@ref(fig:partresplot2)). Additionally, student populations (`CurrentlyStudying`) did not affect electoral party preference in any election (not shown).

@@

Larger migrant populations from Asia, the Middle East, South-Eastern Europe, the United Kingdom and elsewhere, were either associated with Labor support, or have no effect. Of these areas, only South-Eastern European populations significantly affected party preference, with larger populations associating with Labor in 2013 and 2016 (row 2, Figure \@ref(fig:partresplot2)). Speaking other languages (aside from English) however, appears to have had a far stronger effect, as observed through the `OtherLanguage` variable. Electorates with more diverse speech were associated with higher support for the Liberal party from 2004 onwards, with this effect being significant in 2013 and 2016 (see row 7, Figure \@ref(fig:partresplot1)). Furthermore, none of the variables relating to religious beliefs aside from Christianity had a material effect in any election (this includes the Buddhist, Muslim, Jewish, non-religious and Indigenous Australian populations). The association between Christian populations (`Christianity`) and the Liberal party steadily increased over the years, becoming positive and significant in 2016.

@@

In 2001, 2004 and 2016, higher proportions of people that recently (in the past five years) moved house (`DiffAddress`) increased electoral support for the Liberal party (see row 5 in Figure \@ref(fig:partresplot1)). This is somewhat surprising as one might expect house ownership and rental prices to be linked to two-party preference, rather than household mobility (`PropertyOwned` and `RentLoan` were not significant in any election).

Higher proportions of single parent households were associated with Labor support in all elections (albeit insignificant in 2004, see row 6 in Figure \@ref(fig:partresplot2)), whereas the electoral family and household sizes (via the `FamHouseSize` variable) did not appear to be associated with either party.

@@

From 2007 onwards, both marriages (`Married`) and de facto relationships (`DeFacto`) were found to be strong predictors of the two-party preferred vote in favour of the Labor party. In 2001 and 2004 neither of these variables were significant (see rows 4 and 5 in Figure \@ref(fig:partresplot2)).

@@

It is often hypothesised that states have a systematic bias towards one of the two major parties. Boxplots of residuals grouped by state (Figure \@ref(fig:resstate)) show that the data reflects this to only a limited extent. Tasmania and the Australian Capital Territory appeared to have a bias towards Labor, whereas the South Australia and the Northern Territory tend towards voting Liberal. However, there were relatively few electorates in each of these states (five, two, eleven and two respectively), so this apparent result may be due to incumbent effects rather than an actual state-specific bias.

@@

The incumbent party appeared to have a distinct advantage at the next election. The boxplots in Figure \@ref(fig:resincumbent) show that if either of the Labor or Liberal parties won the seat at the previous election, the electorate is likely to vote in their favour, over and above any socio-demographic effects---this effect has not been captured by the model.

@@

Three robustness checks were conducted to confirm model stability. First, a model for each election was re-estimated using only the variables that were found to be significant in at least one of the six elections. The estimated coefficients of the variables in the re-estimated models all fell within their respective 95% confidence intervals from the full models. The second check involved the ten largest pairwise correlations. For each pair, a model for each election was re-estimated omitting one of the two variables. It was found that for each of these pairs, the estimated effect of the remaining variable in the reduced model was within the 95% confidence interval from the full model. The final check was a visual exploration of different variable projections using a tour [@tourr] for each election. No definitive signs of multicollinearity were observed, and as expected (given the nature of spatial data), there was some clumping of electorates for certain projections.

@@

Based on the distribution of the Cook's distance values and the distribution of hat values, a Cook's distance greater than $0.1$ was considered to be influential, and a hat value greater than $0.5$ was considered to have high leverage. Electorates fitting these criteria were flagged and investigated to examine the characteristics driving these values.

The electorate of Sydney (NSW) had a large Cook's distance and high leverage from 2001 to 2007, due to its diverse population (languages, birthplace and religion), high density of young adults (20 to 34 years old), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It remained a strong supporter of the Labor party and the extent of this support is underpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth was found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup---implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections since, the model underpredicts Wentworth's Liberal support.

Lingiari, an electorate making up almost all of the Northern Territory, had consistently high leverage (all years) and was an outlier in all but the 2013 election due to its large Indigenous population, low rates of property ownership and few workers in management or administrative jobs. Fowler (NSW) had a diverse population with a high proportion of migrants, many Buddhists and Muslims, as well as a high proportion of single parent households. These characteristics explain its high leverage in 2001, 2004, 2010 and 2013, and its strong Labor support made it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance were Canberra (ACT) and Durack (WA) in 2013, and Solomon (NT) in 2016.

All of the electorates examined were not unduly influential in the model and therefore no action was required.

@@

This paper explored the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and censuses. As a census did not always occur in the same year as an election, census data for the 2004--2013 elections was generated by employing a method of spatio-temporal imputation. This imputed electoral socio-demographics for the electoral boundaries in place at the time of the election---an approach that is distinctly different from previous work on modelling election outcomes, where census and election data is typically joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables were standardised (to adjust for changing variable scales) and subsets of variables (representing similar information) were combined into factors, resulting in a reduced predictor set. A spatial error model was then estimated for each election, accounting for the inherent spatial structure of the data.

Across the past six elections, most of the socio-demographics driving the electoral two-party preferred vote were found to remain steady, whilst a few (typically weaker) effects varied over time. Industry and type of work were particularly influential. Energy-related and manufacturing/construction jobs, as well as administrative roles and jobs in education and social services were strongly linked with the Liberal party in all elections. Incomes had a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shifted from a weak association with Labor to a significant Liberal effect over the years, and higher education levels were associated with Labor from 2007 (although marginally insignificant in 2010). Electorates with large populations 20 to 34 years were strongly associated with Labor, whilst the 35 to 54 year old bracket also increased Labor support, but to a lesser extent. It was also found that birthplace diversity slightly favoured Labor, relationships (both marriages and de facto relationships) aligned with Labor preference from 2010 onwards, and the influence of Christian populations trended towards Liberal support whilst other religions had negligible effects. Family and household sizes had minimal influence, although electorates with more single parent households were linked with Labor support. Furthermore, the spatial effects were found to be positive in all elections and significant in 2001, 2004 and 2016, meaning that other characteristics that neighbours had in common  (outside of the variables in the model) appeared to be influential in those years.
